Developing a Bubble Chamber Particle Discriminator Using Semi-Supervised Learning

### 2019-02-26 documentation/report.tex
Fixed paper

\usepackage[margin=1in]{geometry}
\usepackage{amsmath}
\usepackage{biblatex}
\usepackage{parskip}
\usepackage{graphicx}
\usepackage{float}
\usepackage{fancyhdr}
\usepackage{lastpage}
\usepackage{titlesec}
\usepackage{enumitem}
\usepackage{subfig}
\usepackage{hyperref}
\usepackage{soul}

@@

\titleformat*{\section}{\large\bfseries}
\titleformat*{\subsection}{\normalsize\bfseries}
\titleformat*{\subsubsection}{\normalsize\bfseries}

@@

\begin{center}
    \begin{Large}
        Improving Particle Classification in WIMP Dark Matter Detection Experiments Using Neural Networks
    \end{Large}

    Brendon Matusch
\end{center}

@@

In all experiments for detection of WIMP dark matter, it is essential to develop a classifier that can distinguish WIMP events from background radiation, in particular alpha particles. Manually developing such a classifier necessitates time-consuming physical modeling.

@@

Previously in PICO-60, a conventional classifier was developed and is thought to be 100\% accurate. However, machine learning is also applicable because it is quicker to develop and re-calibrate. A supervised neural network was used, reaching a mean of 80.2\% accuracy.

In DEAP-3600, a simulation was developed prior to my work, because it is impractical to collect large amounts of calibration data. A conventional classifier removes 99.6\% of alpha background radiation, while also (undesirably) removing 91.0\% of simulated WIMP events.

My objective was to develop novel machine learning algorithms that are more accurate, and I succeeded at this goal! I composed a 28-page research whitepaper \cite{me} about my work on PICO-60, which has been reviewed and approved by the PICO collaboration and pre-published at \url{https://arxiv.org/abs/1811.11308}. It is currently undergoing peer review for eventual publication in the journal \textit{Computer Physics Communications}.

\ul{Please note that the entire PICO collaboration are listed on the research whitepaper because of their contributions to the original PICO-60 dark matter experiment. They did not contribute to this study; I completed and documented it independently.}

@@

For PICO-60, I developed and compared two sets of classification algorithms: supervised learning, and semi-supervised learning.

@@

I experimented further with supervised learning, exploring the following data formats to learn which was the best-performing solution:

\begin{enumerate}
    \item An 8-band Fourier transform of audio recorded in the bubble chamber. I applied a dense neural network with dropout \cite{dropout} and L2 regularization (which allowed for less overfitting to impure data).
    \item A full-resolution Fourier transform of the audio (all 50,001 data points). I once again applied a dense neural network with the Adam \cite{adam} optimizer.
    \item A raw audio waveform with a very deep 1D convolutional neural network (CNN) inspired by Dai et al.\ \cite{verydeepconvnets}.
    \item Images captured by cameras in the detector. I trained a 2D CNN on these, to learn whether they contain any relevant information.
\end{enumerate}

\begin{figure}[ht]
    \centering
    \subfloat{\fbox{\includegraphics[width=0.8\textwidth]{audio_cropped}}}
    \qquad
    \subfloat{\fbox{\includegraphics[width=0.45\textwidth]{image_grid}}}
    \caption{Examples of the audio waveform and Fourier transform (above) and the images captured by cameras (below).}
\end{figure}

@@

To even better handle impurities, I developed two novel semi-supervised learning algorithms. The impure alpha/WIMP-like labels were removed from a large portion of the training data, and my algorithms learned more accurate labels for this data.

@@

In DEAP-3600, the key challenge is the aforementioned unusual detector format: a sphere (with an opening on the top) tiled with a hexagonal lattice of PMTs. I approached this problem in three different ways:

\begin{enumerate}
    \item I tried simply inputting photon counts from each PMT into a multi-layer perceptron.

    \item I developed and applied a Mercator-like cylindrical projection, which maps a sphere onto a rectangle. Cubic spline interpolation was used to map latitude lines on the sphere (which vary in length) to equal-length rows on the rectangular image.

    \item I developed a new type of CNN, called a topological CNN. Rather than a 2D image of square pixels, kernels convolve over an arbitrary topology.

@@

\begin{figure}[ht]
    \centering
    \fbox{\includegraphics[width=0.45\textwidth]{map_projection}}
    \qquad
    \fbox{\includegraphics[width=0.45\textwidth]{topological}}
    \caption{Diagrams of the cylindrical projection (left) and topological CNN (right) systems.}
\end{figure}

@@

Models were optimized using grid searches, in which every configuration within a hyperparameter space is trained and evaluated. All models were trained on a training set, and a separate randomly selected validation set was used to select the best-performing technique and corresponding hyperparameters.

@@

As seen in Figure \ref{pico_final_results}, the banded FFT format was the most effective of the supervised learning models (including the previous study), reaching 97.3\% accuracy on validation data. The image data was shown to contain no meaningful information, permitting only 63.0\% accuracy.

Semi-supervised learning produced better accuracy than supervised learning. Gravitational differentiation was the best overall model, reaching 99.7\% accuracy on validation data and 98.3\% on test data. This confirms my hypothesis; it is indeed beneficial to train on the neural network's most confident predictions.

\begin{figure}[ht]
    \centering
    \fbox{\includegraphics[width=0.85\textwidth]{pico_final_results}}
    \caption{\label{pico_final_results} Validation accuracy, precision, recall, and loss for each technique in PICO-60. Error bars represent the 8th and 92nd percentiles.}
\end{figure}

@@

The three algorithms I developed were evaluated based on their reduction in the rate of false positives (simulated WIMP events misidentified as neck alphas) compared to the 91.0\% result achieved with a conventional classifier. Only models with at least 99.6\% neck alpha removal (in line with the conventional classifier) were considered.

Both the dense neural network and topological CNN produced a higher rate of false positives than the conventional classifier, meaning that they misclassified more WIMPs as alpha particles. However, as seen in Figure \ref{deap_final_results}, the cylindrical projection method \textit{reduced} the false positive rate from 91.0\% to 75.7\%.

This successful network model was tested on the limited amount of available real-world data. It removed 73.9\% of WIMP-like (neutron calibration) events, which is in line with the 75.7\% obtained on simulated data.

However, only 91.2\% of background events (expected to be neck alphas) were removed, quite far off the 99.6\% reached on simulated data. This indicates one of three possibilities:

@@

\begin{figure}[ht]
    \centering
    \fbox{\includegraphics[width=0.5\textwidth]{deap_final_results}}
    \qquad
    \begin{tabular}[b]{|l|l|l|}
        \hline
        Method & Backgnd. & WIMP \\
        \hline
        Conventional & 99.6\% & 91.0\% \\
        \hline
        Dense NN & 100\% & 100\% \\
        \hline
        Cylindrical & 99.9\% & 76.7\% \\
        \hline
        Topological & 99.9\% & 93.0\% \\
        \hline
        Cyl Sim Test & 99.6\% & 75.7\% \\
        \hline
        Cyl Real Test & 91.2\% & 73.9\% \\
        \hline
    \end{tabular}
    \caption{\label{deap_final_results} Background and WIMP removal rates for the highest-accuracy hyperparameter configuration of each technique in DEAP-3600.}
\end{figure}

>>>>

\usepackage[margin=1in]{geometry}
\usepackage{amsmath}
\usepackage{biblatex}
\usepackage{parskip}
\usepackage{graphicx}
\usepackage{float}
\usepackage{fancyhdr}
\usepackage{lastpage}
\usepackage{titlesec}
\usepackage{enumitem}
\usepackage{subfig}
\usepackage{hyperref}
\usepackage{soul}
\usepackage{color}
\usepackage{colortbl}

@@

\titleformat*{\section}{\normalsize\bfseries}
\titleformat*{\subsection}{\normalsize\bfseries}
\titleformat*{\subsubsection}{\normalsize\bfseries}

@@

\begin{center}
    \begin{large}
        Improving Particle Classification in WIMP Dark Matter Detection Experiments Using Neural Networks -- Brendon Matusch
    \end{large}
\end{center}

@@

In all experiments for detection of WIMP dark matter, it is essential to develop a classifier that can distinguish potential WIMP events from background radiation, in particular alpha particles. Manually developing a classifier necessitates time-consuming physical modeling.

@@

Previously in PICO-60, a conventional classifier was developed and is thought to be 100\% accurate. A supervised neural network (NN) was also previously developed, achieving a mean accuracy of 80.2\% -- not usable as a practical replacement for conventional methods.

In DEAP-3600, a physical simulation was developed (prior to my work) to provide calibration data; physical limitations make it impractical to collect large amounts of calibration data. In the simulation, the conventional classifier removes 99.6\% of alpha background radiation, while also (undesirably) removing 91.0\% of WIMP events.

My objective was to develop accurate machine learning algorithms that are more accurate, and I succeeded at this goal! I composed a 28-page research whitepaper \cite{me} about my work on PICO-60, which has been reviewed and approved by the PICO collaboration and pre-published at \url{https://arxiv.org/abs/1811.11308}. It is currently undergoing peer review for eventual publication in the journal \textit{Computer Physics Communications}.

\ul{All PICO researchers are listed on my paper for their work on the original PICO-60 experiment. They did not contribute to this study; I completed and documented it independently.}

@@

For PICO-60, my intent was to develop a highly accurate classification algorithm that is intolerant to substantial impurities in calibration data. I developed and compared two sets of classification algorithms: supervised learning, and semi-supervised learning.

@@

I experimented with supervised learning, exploring the following data formats to determine which facilitates the best-performing solution:

\begin{enumerate}
    \item An 8-band Fourier transform of audio recorded in the detector. I applied a dense NN with dropout \cite{dropout} and L2 regularization (which reduce overfitting to impure data).
    \item A full-resolution Fourier transform of the audio (all 50,001 data points). I once again applied a dense neural network with the Adam \cite{adam} optimizer.
    \item A raw audio waveform, with a very deep 1D convolutional neural network (CNN) inspired by Dai et al.\ \cite{verydeepconvnets}.
    \item Event images, with a 2D CNN (to determine whether they contain useful information).
\end{enumerate}

\begin{figure}[ht]
    \centering
    \subfloat{\fbox{\includegraphics[width=0.95\textwidth]{audio_cropped}}}
    \qquad
    \subfloat{\fbox{\includegraphics[width=0.35\textwidth]{image_grid}}}
    \caption{Examples of the audio and image data.}
\end{figure}

@@

To better handle impurities, I developed two novel semi-supervised learning techniques. The concept of both algorithms is to iteratively generate more accurate labels for a set of inaccurately labeled training data, permitting less overfitting and greater validation accuracy than simple regularization would allow.

@@

In DEAP-3600, the key challenge is the aforementioned unusual detector format: a sphere (with an opening on the top) tiled with a hexagonal lattice of PMTs. The major source of background radiation is ``neck alphas'', which strike atoms in the gas filling the neck above the spherical detector body. These must be eliminated from the experimental data set.

I approached this problem in three different ways, listed below. Techniques \ref{map_entry} and \ref{topological_entry} (Figure \ref{deap_methods}) take advantage of geometric and spatial patterns in PMT activations associated with neck alpha events, by extending the CNN concept for use in spherical and arbitrary topologies.

\begin{enumerate}
    \item I inputted photon counts from each PMT into a multi-layer perceptron.

    \item I developed and applied a Mercator-like cylindrical projection, which maps a sphere onto a rectangle. Cubic spline interpolation was used to map latitude lines on the sphere (which vary in length) to equal-length rows on the rectangular image. Once again, photon counts were used as input data. \label{map_entry}

    \item I developed a new type of CNN, called a topological CNN. Rather than a 2D image of square pixels, kernels convolve over an arbitrary topology. \label{topological_entry}

@@

\begin{figure}[ht]
    \centering
    \fbox{\includegraphics[width=0.425\textwidth]{map_projection}}
    \qquad
    \fbox{\includegraphics[width=0.475\textwidth]{topological}}
    \caption{\label{deap_methods} Diagrams of the cylindrical projection (left) and topological CNN (right) systems.}
\end{figure}

@@

Models were optimized using grid searches, in which every configuration within a hyperparameter space is trained and evaluated. The best-performing technique and hyperparameters were selected based on repeated random sub-sampling of validation sets.

@@

As seen in Figure \ref{pico_final_results}, the banded Fourier transform format was the most effective of the supervised learning models (including the previous study), achieving 97.3\% validation accuracy. The image data was shown to contain no useful information, reaching only 63.0\% accuracy.

Semi-supervised learning produced better accuracy than supervised learning. Gravitational differentiation was the best overall model, achieving 99.7\% accuracy on validation data and 98.3\% on test data. This confirms my hypothesis; it is indeed beneficial to train on the neural network's most confident predictions.

\begin{figure}[ht]
    \centering
    \fbox{\includegraphics[width=0.95\textwidth]{pico_final_results}}
    \caption{\label{pico_final_results} Validation accuracy, precision, recall, and loss for each technique in PICO-60. Error bars represent the 8th and 92nd percentiles.}
\end{figure}

@@

The three algorithms I developed were evaluated based on:

\begin{itemize}
    \item Identification of neck alphas with 99.6\% precision (same as the conventional classifier).
    \item Reduction in the rate of false positives (simulated WIMP events misidentified as neck alphas) compared to the 91.0\% result achieved with a conventional classifier.
\end{itemize}

All techniques were validated on simulated data. Both the dense neural network and topological CNN techniques were unsuccessful; they produced higher false positive rates than the conventional classifier, meaning that they misclassified more WIMPs as alpha particles.

However, as seen in Figure \ref{deap_final_results}, the cylindrical projection method retained a 99.6\% neck alpha identification rate, and also successfully \textit{reduced} the false positive rate from 91.0\% to 75.7\%.

This successful network model was further tested on the limited amount of available real-world experimental data. It removed 73.9\% of WIMP-like (neutron calibration) events, which is in line with the 75.7\% obtained on simulated data.

However, only 91.2\% of background events (expected to be neck alphas) were removed, below the 99.6\% reached on simulated data. This indicates one of three possibilities:

@@

\begin{figure}[ht]
    \centering
    \definecolor{lightcyan}{rgb}{0.8, 1, 1}
    \begin{tabular}[b]{|l|l|l|l|}
        \hline
        \rowcolor{lightcyan}
        Technique & Data Source & Background & False Positive Rate \\
        \rowcolor{lightcyan}
        & & Removal Rate & (WIMP Misclassifications) \\
        \hline
        Conventional Classifier & Simulation & 99.6\% & 91.0\% \\
        \hline
        Dense Neural Network & Simulation & 100\% & 100\% \\
        \hline
        Cylindrical Projection & Simulation & 99.9\% & 76.7\% \\
        \hline
        Topological CNN & Simulation & 99.9\% & 93.0\% \\
        \hline
        Cylindrical Projection & Simulation (Test) & 99.6\% & 75.7\% \\
        \hline
        Cylindrical Projection & Real-World & 91.2\% & 73.9\% \\
        \hline
    \end{tabular}
    \caption{\label{deap_final_results} Average background and WIMP/neutron removal rates for the highest-accuracy hyperparameter configuration of each technique in DEAP-3600.}
\end{figure}
